Leaf traits drive differences in biomass partitioning among major plant functional types

### 2014-12-17 manuscript.Rnw
Tidied references.bib

\affil[1]{1 Hawkesbury Insitute for the Environment, University of Western Sydney, Locked Bag 1797, Penrith, NSW 2751, Australia}
\affil[2]{Biological Sciences, Macquarie University, North Ryde, NSW 2109, Australia}
\affil[3]{See Supporting Information}

@@

\textbf{
The partitioning of biomass into short-lived leaves and long-lived stems is one of the most uncertain  and influential components of  global vegetation models (GVMs), affecting the lifetime of carbon in ecosystems \cite{friend_carbon_2014}, predictions of global carbon storage \cite{ise_comparison_2010}, and the CO\textsubscript{2} response of net primary production \cite{dekauwe2014}. Although GVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical data have been lacking. Using a new global database on biomass partitioning from over 175 studies, we here show a strong effect of PFT on partitioning between stems and leaves. Contrary to prior expectation, we show that major woody PFTs do not differ in the amount of leaf area supported, but rather the amount of leaf mass per unit above ground biomass. Moreover, we find that these differences relate to systematic shifts in leaf mass per area, a leading trait within the leaf economics spectrum. Combined, these results uncover fundamental rules in how plants are constructed and allow for systematic benchmarking of biomass partitioning routines in current GVMs.
}

@@

\item GVMs (including dynamic GVMs, which model the distribution of vegetation types, and those who use a prescribed vegetation distributions) are used to predict the trajectory of the global vegetation carbon sink in a changing climate. Biomass partitioning (related to, but not to be confused with allocation of net primary production) is important because short-lived leaves re-enter the decomposition cycle quickly, releasing much of its carbon back to the atmosphere. Long-lived stems act as a very long carbon storage term (Luyssaert). As a result, partitioning affects the overall lifetime of carbon in the ecosystem, which has been shown to be an important determinant of overall carbon storage projections(Friend). 

@@

\item What is known is that gymnosperms have more leaf biomass per unit aboveground biomass (the 'leaf mass fraction', LMF) than angiosperms (Enquist etc.). But it is not even clear whether this is due to the angiosperm/evergreen divergence, or related to inherent differences in leaf life span. If the latter, we would expect evergreen angiosperms to have more leaf biomass than deciduous angiosperms. 

@@

\item We then found that LMF was proportional to the leaf construction cost, as measured by leaf mass per area (Fig 2b). \item The consequence of this is that the leaf area ratio (the amount of leaf area supported per unit aboveground biomass), was invariant between PFTs, because $LAR = LMF / LMA$ (Fig 2c and S3). We thus supported the view that PFTs differ in LMF, but for the first time demonstrated that they converge in the amount of leaf area supported.
\item Next, we split LMF into two components, $LMF = M_F/M_T = M_F/A_S * A_S/MT$, where $M_F/A_S$ is the ratio of leaf biomass to stem basal area (see Methods S1), and $A_S/M_T$ the ratio of stem basal area to total aboveground biomass. We found that the first term showed identical PFT patterns as LMF (Fig 2d and e), $M_F/A_S$ was highly variable between PFT and proportional to LMA, so that $A_F/A_S$ is invariant between PFTs. 
\item This split is also convenient because the first term is nearly independent of plant height, allowing a more straightforward analysis of climate effects (not having to correct for height every time). The second term was found to be nearly unaffected by PFT (S4), but strongly height-dependent. It thus absorbs variation related to height and aboveground plant geometry, and is consistent with previous findings that this ratio is fairly conserved between plant species (Chaves).

@@

\item We found very modest effects of climate on biomass partitioning, somewhat inconsistent with Reich et al 2014. It is consistent with the idea that LMA is more flexible than biomass partitioning (Poorter). We argue that climate directly or indirectly affects LMA, which then leads to differences in LMF, because LAR is less variable. Plants seem to target a certain total leaf area display per unit aboveground biomass investment, but we do stress that there is still much variation unaccounted for (Figure S histograms), and that other studies certainly have uncovered small-scale effects of climate on leaf area to sapwood area (pine studies etc), an important component of $A_F/A_S$. Our results hold at the PFT level, which is relevant for DGVMs.
\item What consequences for DGVMs? There is a multitude of allocation models (de Kauwe). Together with assumptions on leaf and stem turnover rates, this gives estimates of biomass partitioning (S2 methods). So far it is common to choose between different models largely based on philosophical arguments (Franklin 2012). We argue that relationships such as in Fig 2a should be used to benchmark and constrain the various allocation models. The data we use here is freely available (Falster), which should aid in the models being tested, with the hope that they perhaps converge.

@@

Major plant functional types differ widely in leaf mass per area (LMA, the ratio of leaf mass to area, an index of leaf thickness) (\cite{poorter_causes_2009}), which can be expected to affect biomass partitioning in several ways. First, LMA correlates with leaf lifespan. Longer lived leaves would lead to higher plant leaf mass ($M_F$) as long as annual production rates are similar (\cite{enquist_global_2002}). Another interpretation is that plants may target a certain leaf area display given a certain biomass investment in supporting tissues, because light interception and photosynthesis correlate with the exchange area of the plant.

It has been reported numerous times that gymnosperms (with higher LMA) have higher leaf mass fraction than angiosperms (with lower LMA) \cite{enquist_global_2002, cheng_relationships_2007, mccarthy_consistency_2007, poorter_biomass_2012, sardans_tree_2013}, although a recent study reported no difference (\cite{wolf_allometric_2010}). No study to date has been able to further tease apart PFT differences, or test the potentially confounding effects of climate.

We can further decompose LMF into two components representing different aspects of plant function and structure (see Methods). The ratio of leaf biomass ($M_F$) to stem basal area ($A_S$) relates to hydraulic supply, and the ratio of total aboveground biomass ($M_T$) and $A_S$ is a function of plant shape (stem taper, plant height at given stem diameter), and wood density \cite{chave_tree_2005}. We found that ratio of aboveground biomass to basal stem area ($M_T/A_S$) was nearly constant between PFTs (Fig.~\ref{fig:figureS4}), with deciduous angiosperms having slightly higher $M_T$ at a given $A_S$. However we found very large differences between PFTs in the ratio $M_F/A_S$(Fig.~\ref{fig:figure2}).

@@

Unlike \cite{reich_temperature_2014}, we found only very weak effects of climate on biomass relationships. The variance explained by MAT was very small, especially compared to the large difference between PFTs.

Leaf mass per area generally shows a more plastic response to climate than biomass partitioning  \cite{poorter_biomass_2012}.

@@

There is no shortage of theoretical developments in understanding biomass allocation (\cite{cannell_carbon_1994,franklin_modeling_2012,enquist_land_2012}). In particular, theoretical work focusing on optimal partitioning theory is promising (e.g. \cite{valentine_modeling_2012}) but is yet to lead to operational models for dynamic global vegetation models.

We found no effects of MAP on biomass partitioning, and only very weak effects of MAT on LMF, when analyzed in addition to PFT. This means that climate-related differences are captured by a shift in PFT (see also Fig.~\ref{fig:figure1}, showing PFTs occupy different parts of the climate space). Our results contrast with those of \cite{reich_temperature_2014} but they studied whole-stand biomass, not individual plant biomass. So perhaps this means the climate effects are on stand density not individual plant effects. No effect of climate is good news for models that parameterize the globe by PFT, as this seems sufficient, based on our dataset.

@@

We used the recently compiled Biomass And Allometry Database (BAAD) (Falster et al., in review), including a total of \Sexpr{nrow(baad_all$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), aboveground woody biomass ($M_W$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (ca. 1.3m) ($A_{Sbh}$).  For each analysis, we used different subsets of the data. Sample sizes by PFT are summarized in Tables~\ref{table:count1}-\ref{table:count4} We exluded glasshouse and common garden studies, including only field-grown woody plants (including natural vegetation, unmanaged and managed plantations). We considered three PFTs : evergreen angiosperms, evergreen gymnosperms, and decidous angiosperms. We excluded  deciduous gymnosperms because few data were available. All locations were further separated into boreal (including sub-boreal), temperate, and tropical biomes.

For the aboveground biomass pools, we calculated the leaf mass fraction ($M_F / M_T$, where $M_T$ is total aboveground biomass) and leaf area ratio ($A_F / M_T$). These variables are related by,

@@

We further decomposed LMF and LAR by considering the ratio of leaf mass (or area) to plant basal area, and the ratio of aboveground biomass to plant basal area. For LMF,

@@

And similarly for LAR, by replacing $M_F$ with $A_F$. We were unable to further decompose $A_S/M_T$ into wood volume and wood density components (cf. \cite{chave_tree_2005}) because wood density data were available only for a small subset of data.

We only used LMA directly estimated for the harvested plants (typically for a subsample of leaves). For conifers, leaf area was converted to half-total surface area, which is most appropriate for comparison to flat leaves in terms of light interception  \cite{lang_application_1991, chen_defining_1992}.

@@

We used a generalized additive model (GAM) to visualize the relationships between biomass and plant size variables. In addition, we corrected for plant size by fitting linear mixed-effects models (with species within study as the random effect) using plant height (linear and squared terms), and calculating least-square means with the 'lsmeans' package in R. All variables were log-transformed prior to analysis. Variance partitioning with marginal (fixed effects) and conditional (fixed and random effects) $R^2$ for linear mixed-effects models as reported by \cite{nakagawa_general_2013}.

@@

Figure 1 coutresy of Lana Heydon, IAN Image Library http://ian.umces.edu/imagelibrary.

@@

R.A.D and D.S.F. conceived the ideas and methods, and co-wrote the paper. R.A.D. analysed data. BAAD contributors shared data.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{Figure2.pdf}
    \caption{\textbf{Dominant woody PFTs differ in leaf mass fraction due to underlying differences in leaf construction cost. } (a) Leaf mass fraction (LMF = leaf mass / aboveground biomass) by PFT. Each symbol is an individual plant. Lines are generalized additive model fits. (b) and (c) Leaf mass fraction (LMF) and leaf area ratio (LAR) at the average plant height in the dataset, estimated from panel (a). (d) Average leaf mass per unit basal stem area, and (e) leaf area per unit basal stem area for the three major PFTs confirm that the between-PFT variation in leaf mass fraction is due to leaf mass per unit basal area. Error bars are 95\% confidence intervals. Letters denote significant differences (at $\alpha = 0.05$).}
    \label{fig:figure2}
\end{figure}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{FigureS4.pdf}
    \caption{\textbf{Relationship between aboveground biomass and basal stem area, each point is an individual plant.} Lines are spline fits. The right panel shows least-square means of the ratio of aboveground biomass to basal stem area, estimated from a linear mixed-effects model with basal stem area (linear and squared terms) and PFT. }
    \label{fig:figureS4}
\end{figure}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{FigureS6.pdf}
    \caption{\textbf{Plant leaf mass as a function of basal stem area (left panel), and leaf area as a function of basal stem area (right panel).} Individual plant data were averaged by species/study (the same species occuring in multiple datasets thus appears multiple times). Lines are linear regressions with 95\% confidence intervals. Both regressions were significant, but $R^2$ was 0.28 in the left panel, and only 0.02 in the right panel.}
  \label{fig:figureS6}
\end{figure}

>>>>

\affil[1]{Hawkesbury Institute for the Environment, University of Western Sydney, Locked Bag 1797, Penrith, NSW 2751, Australia}
\affil[2]{Biological Sciences, Macquarie University, North Ryde, NSW 2109, Australia}
\affil[3]{See Supporting Information}

@@

\textbf{
The partitioning of biomass into short-lived leaves and long-lived stems is one of the most uncertain  and influential components of  global vegetation models (GVMs), affecting the lifetime of carbon in ecosystems \cite{friend2014}, predictions of global carbon storage \cite{ise2010}, and the CO\textsubscript{2} response of net primary production \cite{dekauwe2014}. Although GVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical data have been lacking. Using a new global database on biomass partitioning from over 175 studies, we here show a strong effect of PFT on partitioning between stems and leaves. Contrary to prior expectation, we show that major woody PFTs do not differ in the amount of leaf area supported, but rather the amount of leaf mass per unit above ground biomass. Moreover, we find that these differences relate to systematic shifts in leaf mass per area, a leading trait within the leaf economics spectrum. Combined, these results uncover fundamental rules in how plants are constructed and allow for systematic benchmarking of biomass partitioning routines in current GVMs.
}

@@

\item GVMs (including dynamic GVMs, which model the distribution of vegetation types, and those who use a prescribed vegetation distributions) are used to predict the trajectory of the global vegetation carbon sink in a changing climate. Biomass partitioning (related to, but not to be confused with allocation of net primary production) is important because short-lived leaves re-enter the decomposition cycle quickly, releasing much of its carbon back to the atmosphere. Long-lived stems act as a very long carbon storage term \cite{luyssaert2008}. As a result, partitioning affects the overall lifetime of carbon in the ecosystem, which has been shown to be an important determinant of overall carbon storage projections \cite{friend2014}.

@@

\item What is known is that gymnosperms have more leaf biomass per unit above-ground biomass (the 'leaf mass fraction', LMF) than angiosperms \cite{oneilldeangelis1981,enquistniklas2002}. But it is not even clear whether this is due to the angiosperm/evergreen divergence, or related to inherent differences in leaf life span. If the latter, we would expect evergreen angiosperms to have more leaf biomass than deciduous angiosperms. 

@@

\item We then found that LMF was proportional to the leaf construction cost, as measured by leaf mass per area (Fig 2b). \item The consequence of this is that the leaf area ratio (the amount of leaf area supported per unit above-ground biomass), was invariant between PFTs, because $LAR = LMF / LMA$ (Fig 2c and S3). We thus supported the view that PFTs differ in LMF, but for the first time demonstrated that they converge in the amount of leaf area supported.
\item Next, we split LMF into two components, $LMF = M_F/M_T = M_F/A_S * A_S/MT$, where $M_F/A_S$ is the ratio of leaf biomass to stem basal area (see Methods S1), and $A_S/M_T$ the ratio of stem basal area to total above-ground biomass. We found that the first term showed identical PFT patterns as LMF (Fig 2d and e), $M_F/A_S$ was highly variable between PFT and proportional to LMA, so that $A_F/A_S$ is invariant between PFTs. 
\item This split is also convenient because the first term is nearly independent of plant height, allowing a more straightforward analysis of climate effects (not having to correct for height every time). The second term was found to be nearly unaffected by PFT (S4), but strongly height-dependent. It thus absorbs variation related to height and above-ground plant geometry, and is consistent with previous findings that this ratio is fairly conserved between plant species (Chaves).

@@

\item We found very modest effects of climate on biomass partitioning, somewhat inconsistent with Reich et al 2014. It is consistent with the idea that LMA is more flexible than biomass partitioning (Poorter). We argue that climate directly or indirectly affects LMA, which then leads to differences in LMF, because LAR is less variable. Plants seem to target a certain total leaf area display per unit above-ground biomass investment, but we do stress that there is still much variation unaccounted for (Figure S histograms), and that other studies certainly have uncovered small-scale effects of climate on leaf area to sapwood area (pine studies etc), an important component of $A_F/A_S$. Our results hold at the PFT level, which is relevant for DGVMs.
\item What consequences for DGVMs? There is a multitude of allocation models (de Kauwe). Together with assumptions on leaf and stem turnover rates, this gives estimates of biomass partitioning (S2 methods). So far it is common to choose between different models largely based on philosophical arguments \cite{franklin2012}. We argue that relationships such as in Fig 2a should be used to benchmark and constrain the various allocation models. The data we use here is freely available (Falster), which should aid in the models being tested, with the hope that they perhaps converge.

@@

Major plant functional types differ widely in leaf mass per area (LMA, the ratio of leaf mass to area, an index of leaf thickness) (\cite{poorter2009}), which can be expected to affect biomass partitioning in several ways. First, LMA correlates with leaf lifespan. Longer lived leaves would lead to higher plant leaf mass ($M_F$) as long as annual production rates are similar (\cite{enquistniklas2002}). Another interpretation is that plants may target a certain leaf area display given a certain biomass investment in supporting tissues, because light interception and photosynthesis correlate with the exchange area of the plant.

It has been reported numerous times that gymnosperms (with higher LMA) have higher leaf mass fraction than angiosperms (with lower LMA) \cite{enquistniklas2002, cheng2007a, mccarthyetal2007, poorter2012, sardans2013}, although a recent study reported no difference (\cite{wolf2010}). No study to date has been able to further tease apart PFT differences, or test the potentially confounding effects of climate.

It is customary to estimate total above-ground biomass from stem diameter \cite{jenkins2003,chave2005}.

We can further decompose LMF into two components representing different aspects of plant function and structure (see Methods). The ratio of leaf biomass ($M_F$) to stem basal area ($A_S$) relates to hydraulic supply, and the ratio of total above-ground biomass ($M_T$) and $A_S$ is a function of plant shape (stem taper, plant height at given stem diameter), and wood density \cite{chave2005}. We found that ratio of above-ground biomass to basal stem area ($M_T/A_S$) was nearly constant between PFTs (Fig.~\ref{fig:figureS4}), with deciduous angiosperms having slightly higher $M_T$ at a given $A_S$. However we found very large differences between PFTs in the ratio $M_F/A_S$(Fig.~\ref{fig:figure2}).

@@

Unlike \cite{reich2014}, we found only very weak effects of climate on biomass relationships. The variance explained by MAT was very small, especially compared to the large difference between PFTs.

Leaf mass per area generally shows a more plastic response to climate than biomass partitioning  \cite{poorter2012}.

@@

There is no shortage of theoretical developments in understanding biomass allocation (\cite{cannelldewar1994,franklin2012,enquist2012}). In particular, theoretical work focusing on optimal partitioning theory is promising (e.g. \cite{valentinemakela2012}) but is yet to lead to operational models for dynamic global vegetation models.

We found no effects of MAP on biomass partitioning, and only very weak effects of MAT on LMF, when analyzed in addition to PFT. This means that climate-related differences are captured by a shift in PFT (see also Fig.~\ref{fig:figure1}, showing PFTs occupy different parts of the climate space). Our results contrast with those of \cite{reich2014} but they studied whole-stand biomass, not individual plant biomass. So perhaps this means the climate effects are on stand density not individual plant effects. No effect of climate is good news for models that parameterize the globe by PFT, as this seems sufficient, based on our dataset.

@@

We used the recently compiled Biomass And Allometry Database (BAAD) (Falster et al., in review), including a total of \Sexpr{nrow(baad_all$data)} observations, but here we use only data that include leaf mass ($M_F$), leaf area ($A_F$), above-ground woody biomass ($M_W$), plant height ($H$), and stem area measured at ground level ($A_S$), or at breast height (ca. 1.3m) ($A_{Sbh}$).  For each analysis, we used different subsets of the data. Sample sizes by PFT are summarized in Tables~\ref{table:count1}-\ref{table:count4} We excluded glasshouse and common garden studies, including only field-grown woody plants (including natural vegetation, unmanaged and managed plantations). We considered three PFTs : evergreen angiosperms, evergreen gymnosperms, and deciduous angiosperms. We excluded  deciduous gymnosperms because few data were available. All locations were further separated into boreal (including sub-boreal), temperate, and tropical biomes.

For the above-ground biomass pools, we calculated the leaf mass fraction ($M_F / M_T$, where $M_T$ is total above-ground biomass) and leaf area ratio ($A_F / M_T$). These variables are related by,

@@

We further decomposed LMF and LAR by considering the ratio of leaf mass (or area) to plant basal area, and the ratio of above-ground biomass to plant basal area. For LMF,

@@

And similarly for LAR, by replacing $M_F$ with $A_F$. We were unable to further decompose $A_S/M_T$ into wood volume and wood density components (cf. \cite{chave2005}) because wood density data were available only for a small subset of data.

We only used LMA directly estimated for the harvested plants (typically for a subsample of leaves). For conifers, leaf area was converted to half-total surface area, which is most appropriate for comparison to flat leaves in terms of light interception  \cite{lang1991, chen1992}.

@@

We used a generalized additive model (GAM) to visualize the relationships between biomass and plant size variables. In addition, we corrected for plant size by fitting linear mixed-effects models (with species within study as the random effect) using plant height (linear and squared terms), and calculating least-square means with the 'lsmeans' package in R. All variables were log-transformed prior to analysis. Variance partitioning with marginal (fixed effects) and conditional (fixed and random effects) $R^2$ for linear mixed-effects models as reported by \cite{nakagawa2013}.

@@

Figure 1a uses an image by Lana Heydon, IAN Image Library http://ian.umces.edu/imagelibrary.

@@

R.A.D and D.S.F. conceived the ideas and methods, and co-wrote the paper. R.A.D. analysed data. Co-authors in the BAAD consortium shared data.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{Figure2.pdf}
    \caption{\textbf{Dominant woody PFTs differ in leaf mass fraction due to underlying differences in leaf construction cost. } (a) Leaf mass fraction (LMF = leaf mass / above-ground biomass) by PFT. Each symbol is an individual plant. Lines are generalized additive model fits. (b) and (c) Leaf mass fraction (LMF) and leaf area ratio (LAR) at the average plant height in the dataset, estimated from panel (a). (d) Average leaf mass per unit basal stem area, and (e) leaf area per unit basal stem area for the three major PFTs confirm that the between-PFT variation in leaf mass fraction is due to leaf mass per unit basal area. Error bars are 95\% confidence intervals. Letters denote significant differences (at $\alpha = 0.05$).}
    \label{fig:figure2}
\end{figure}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{FigureS4.pdf}
    \caption{\textbf{Relationship between above-ground biomass and basal stem area, each point is an individual plant.} Lines are spline fits. The right panel shows least-square means of the ratio of above-ground biomass to basal stem area, estimated from a linear mixed-effects model with basal stem area (linear and squared terms) and PFT. }
    \label{fig:figureS4}
\end{figure}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{FigureS6.pdf}
    \caption{\textbf{Plant leaf mass as a function of basal stem area (left panel), and leaf area as a function of basal stem area (right panel).} Individual plant data were averaged by species/study (the same species occurring in multiple datasets thus appears multiple times). Lines are linear regressions with 95\% confidence intervals. Both regressions were significant, but $R^2$ was 0.28 in the left panel, and only 0.02 in the right panel.}
  \label{fig:figureS6}
\end{figure}
